---
title: "Data Visualisation with ggplot2 - Philemon K Langat"
output: html_notebook
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the research question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?


## 2. Reading the Data
```{r}
library("tidyverse") 
```

```{r}
# Load the data below
# --- 
# Dataset url = 
# --- 
# YOUR CODE GOES BELOW
bookinghotel_df=read.csv("/cloud/project/hotel_bookings.csv",header = TRUE, stringsAsFactors = FALSE)
head(bookinghotel_df)
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(bookinghotel_df)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(bookinghotel_df)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample(bookinghotel_df)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(bookinghotel_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(bookinghotel_df)
```


Record your general observations below:

Observation 1- On the sample record - Company is null
Observation 2- The total record for this data (119390-rows and 32 columns)


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
is.na(bookinghotel_df)

```
```{r}
#To find the total missing values in each column we use the `colSums()` function and pass is.na(dt) as the argument and where `dt` is the dataframe.
colSums(is.na(bookinghotel_df))

```

We observe the following from our dataset:

Observation 1- Most of all columns don't have missing data
Observation 2 - This data is good analysis

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
bookinghotel_df[duplicated(bookinghotel_df),]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW

# 
colSums(is.na(bookinghotel_df))
```

We observe the following from our dataset:

Observation 1- There are 31994 duplicate records even though I chose to retain them.

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
# 
rowSums(is.na(bookinghotel_df))

```

We observe the following from our dataset:

Observation 1 -None of the rows data is null
Observation 2- Retained all the data
```{r}
# Let's first install and import the reshape2 package that will provide 
# the melt() function for creating a correlation matrix. 
# After running the following line, once prompted in the console,
# type `Yes` to install it.
install.packages("reshape2")

# loading the package
library(reshape2)
```
```{r}
glimpse(bookinghotel_df)
```

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heat map
# ---
# YOUR CODE GOES BELOW
# Select the desired columns
bookinghotel_df1 <- bookinghotel_df %>%
  na.omit() %>%
  select(is_repeated_guest,children,required_car_parking_spaces,stays_in_week_nights)
#
# Create a correlation matrix
corr_matrix <- cor(bookinghotel_df1, method="s")
head(corr_matrix)
# Create a table with long form
corr_df <- melt(corr_matrix)
head(corr_df)
# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "yellow", 
    high = "grey",
    mid = "black", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
bookinghotel_df1 =select(bookinghotel_df, -c('company'))

glimpse(bookinghotel_df1)

```

From the data variable description we see that the Distribution Channel categorically that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#
bookinghotel_df$agent[is.na(bookinghotel_df$agent)] <- "TO"
sample(bookinghotel_df)
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#bookinghotel_df <-bookinghotel_df %>% select(adults,babies,children,-contains(0))
#bookinghotel_df <- bookinghotel_df %>% filter(!is.null(adults) | adults!=0)
#bookinghotel_df[!(is.na(bookinghotel_df$adults) | bookinghotel_df$adults==0), ]
#bookinghotel_df

bookinghotel_df %>%
  filter(adults>0,children>0,babies>0)


```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
bookinghotel_df$children[is.na(bookinghotel_df$children)] <- round(mean(bookinghotel_df$children, na.rm = TRUE))
bookinghotel_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
colSums(is.na(bookinghotel_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
bookinghotel_df$children <- as.integer(bookinghotel_df$children)
#bookinghotel_df$company <- as.integer(bookinghotel_df$company)
bookinghotel_df$agent <- as.integer(bookinghotel_df$agent)
glimpse(bookinghotel_df)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot
# Create the summary table
bookinghotel_df %>% 
  group_by(hotel) %>%
  summarize(is_canceled = sum(is_canceled))

# Plotting the chart
ggplot(bookinghotel_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "yellow"
  ) +
  labs(
    title = "Total Canceled for Hotel Bookings", 
    x = "Hotels", 
    y = "Total Canceled"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )





```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel


# Create the summary table

bookinghotel_df %>% 
  filter(hotel == "city hotel", hotel== "Resort Hotel")%>%
  group_by(hotel) %>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))

# Plotting the chart
ggplot(bookinghotel_df, aes(x = hotel, y = previous_bookings_not_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "grey"
  ) +
  labs(
    title = "Booking Ration between Resort and City Hotels", 
    x = "Hotels", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )




```
```{r}
# Create summarize data table
bookinghotel_df3 <- bookinghotel_df %>%
  select(arrival_date_year,previous_bookings_not_canceled) %>%
  group_by(arrival_date_year)%>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))
bookinghotel_df3
```


```{r}
# 3. What was the percentage of booking for each year?
# ---

# Plotting and styling
ggplot(data=bookinghotel_df3, aes(x="", y=previous_bookings_not_canceled, fill=as.factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
     aes(label = paste0(round(previous_bookings_not_canceled / sum(previous_bookings_not_canceled) * 100, 1), "%"))
, position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +   
  labs(
    fill = "arrival_date_year", 
    title = "Percentage of Bookings for each (year)"
  ) +
  theme_void()
```
```{r}
# Create summarize data table
bookinghotel_df4 <- bookinghotel_df %>%
  select(arrival_date_month,previous_bookings_not_canceled,hotel) %>%
  group_by(arrival_date_month)%>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))
bookinghotel_df4
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 

# Plotting
# Plotting the chart
ggplot(bookinghotel_df4, aes(x = arrival_date_month, y = previous_bookings_not_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "grey"
  ) +
  labs(
    title = "Busiest Months", 
    x = "Months", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )




```
```{r}
# Create summarize data table

bookinghotel_df5 <- bookinghotel_df %>%
  select(country, adults,children,babies) %>%
  group_by(country)%>%
  summarize(total_guests_per_country = sum(adults+children+babies))%>%
  arrange(desc(total_guests_per_country))
head(bookinghotel_df5,3)
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#

bookinghotel_df5 <- bookinghotel_df %>%
  select(country, adults,children,babies) %>%
  group_by(country)%>%
  summarize(total_guests_per_country = sum(adults,children,babies))%>%
  arrange(desc(total_guests_per_country))
head(bookinghotel_df5,3)



```

```{r}
# 6.a) How long do most people stay in hotels?
 #b) By city and resort? Separate the data by hotel

bookinghotel_df6 <- bookinghotel_df %>%
  select(hotel, stays_in_weekend_nights,stays_in_week_nights) %>%
  group_by(hotel)%>%
  summarize(total_stay_in_hotel = round(mean(stays_in_weekend_nights+stays_in_week_nights)))%>%
  arrange(desc(total_stay_in_hotel))
head(bookinghotel_df6,5)


#  
# ---
# 
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
bookinghotel_df6 <- bookinghotel_df %>%
  select(assigned_room_type,previous_bookings_not_canceled) %>%
  group_by(assigned_room_type)%>%
  summarize(most_booked_accommodation_type = sum(previous_bookings_not_canceled))%>%
  arrange(desc(most_booked_accommodation_type))
head(bookinghotel_df6,30)
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 
```

Our observations:
-

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?-
b). Do we need other data to answer our question?
c). Did we have the right question?


